Internal Gravity Wave Breaking in White Dwarf Binaries

### 2019-11-05 writeups/Su_IGW_break.tex

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\uv}[1]{\hat{\boldsymbol{\mathbf{#1}}}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}
    \DeclareMathOperator*{\erf}{erf}

@@

In practice, it is convenient to introduce the coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$ \citep[e.g.][]{lecoanet_anel}. This both
identically enforces $\rho > 0$ and eliminates the stiff term
$\frac{\bm{\nabla}P}{\rho}$. We also define reduced pressure $\varpi =
\frac{P}{\rho}$. Then, we may rewrite the second two equations of
\autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}}{t} + \bm{\nabla}\varpi + \varpi\bm{\nabla}\Upsilon
            - \frac{\varpi}{H}\uv{z} + g\uv{z} &= 0.
    \end{align}
\end{subequations}
In the new coordinates, hydrostatic equilibrium corresponds to $\Upsilon = 0,
\overline{\varpi} = gH$.

@@

    \item The averaged horizontal momentum flux $F$ is
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def},\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        Thus, indeed $F > 0$ for an upward propagating IGW $c_{g, z} > 0$.
\end{itemize}

@@

The detailed onset of wave breaking has been laid out in a few key papers
\citep{drazin, klostermeyer, winters1994}, where instabilities transfer momentum
out of the IGW through daughter modes into the mean flow of the fluid. When
viscosity is sufficiently weak (see \autoref{ss:disp} for a discussion of this
assumption in WDs), a non-uniform shear flow and eventually a critical layer
develop. The further evolution of the critical layer with the incident IGW is
thought to be responsible for tidal synchronization in stellar binaries
\citep{zahn75,gn89} as well as the quasi-biennial oscillation
\citep{lindzen_qbo}.

The behavior of an incident IGW upon a critical layer was first studied in the
inviscid, linear regime in \citep{booker_bretherton}, which found nearly
complete absorption of the IGW\@. In particular, the incident wave has amplitude
reflection and transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\overline{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R}_A &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T}_A &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the height of the critical
layer $z_c$. In the $\mathrm{Ri} \gg 1$ limit, $\mathcal{R}, \mathcal{T} \ll 1$
and the incident wave is completely absorbed to good approximation.

This result extends to viscous fluids \citep{hazel}, but weakly nonlinear theory
\citep{brown_stewartson} and numerical simulations \citep{winters1994} suggest
that nonlinear effects significantly enhance reflection and transmission. These
studies are not directly applicable to our problem where a continuous train of
IGW are excited.

Instead, we mayanalyze horizontal momentum transfer at the critical layer. Any
incident horizontal momentum flux absorbed by the fluid must manifest as
additional horizontal momentum of the shear flow. Since the mean flow cannot
exceed $\overline{U}_c \equiv \omega/k_x$ the horizontal phase velocity of the
incident wave, the critical layer must instead propagate downwards to
accommodate the incident momentum flux. The total horizontal momentum of the
shear flow obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits \rho(z) \overline{U}(z, t)\;\mathrm{d}z
        + \Delta F = 0,
\end{equation}
where $\Delta F$ is the absorbed horizontal momentum. Assuming $\overline{U}(z >
z_c) = \overline{U}_c, \overline{U}(z < z_c) = 0$, the conservation equation
becomes differential equation for $z_c(t)$
\begin{equation}
    \rho(z_c) \overline{U}_c\pd{z_c}{t} = \Delta F.\label{eq:zc_anal}
\end{equation}
If the entirety of the horizontal momentum flux carried by the IGW wavetrain is
absorbed by the critical layer, then $\Delta F = -F'$ is constant in time.
Finally, for constant $\Delta F$ in time and $\rho \approx \overline{\rho}$,
$z_c(t)$ has analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\overline{\rho}_0\overline{U}_c}{-\Delta F}
        .\label{eq:zc_sol}
\end{equation}

\section{Internal Gravity Waves: Linear Numerical Simulation}\label{s:sim}

Towards numerical simulation of IGW breaking, we first verify agreement with
linear theory at weak forcing amplitudes. We perform direct numerical simulation
using the pseudo-spectral code Dedalus \citep{dedalus}. In \autoref{ss:numerics}
we discuss choices of numerical parameters, in \autoref{ss:damping} we
separately discuss use of a damping zone at the top and bottom of the domain,
and in \autoref{ss:lin_ns} we present the results of our ``linear'' simulations,
where we solve the full nonlinear fluid equations for weak forcing.

@@

We nondimensionalize by taking $H = N = \overline{\rho}_0 = 1$.

In a realistic WD, IGW are excited very deep in the WD and propagate a
substantial number of density scale heights before reaching the breaking zone.
This propagation is very nearly exactly linear until just a few scale heights
below the breaking region. In the interest of computational feasibility, we
restrct the simulation domain to the upper WD fluid, of much smaller vertical
extent than the full physical domain. To further decrease computational cost, we
restrict ourselves to 2D, plane-parallel IGW rather than full 3D stellar
g-modes.

Denote $L_x, L_z$ to be the dimensions of the simulation domain. As the physical
domain extends beyond $L_x, L_z$, we must take care to suppress reflections. We
use periodic boundary conditions in the $x$ direction and damping layers
(described in \autoref{ss:damping}) in the $z$ direction to damp perturbations
incident on the top and bottom of the simulation. We use a Fourier basis in both
the $x$ and $z$ direction. We varied the number of $x, z$ modes (denoted $N_x,
N_z$ respectively), and used $3/2$ dealiasing \citep{boyd}.

The geometry of our simulation domain is fixed by one further parameter: $z_0$
the forcing location. We use $z_0 = 0.2L_z$ to force sufficiently far from the
lower damping zone and permit sufficient room for the upward moving wave to
grow at $\propto e^{z/2H}$. We choose $L_z = 10H$ to give $\sim e^3$ apmlitude
growth between the damping zones. Finally, we want similar grid spacing
$\frac{L_x}{N_x} \sim \frac{L_z}{N_z}$, guided by the intuition that turbulence
is approximately isotropic. For computational savings, we fix $L_x = 4H < L_z$,
a compromise between using fewer $N_x$ modes and using as large of an $L_x$ as
possible (in stars, $L_x \sim 2\pi R$ the circumference). We then fix ratio $N_z
/ N_x = 4$.

@@

The physics of our simulation is fixed by four parameters: $k_{x}, \omega, C,
\nu$. We describe our choices for these parameters below:
\begin{itemize}
    \item $k_{x}$: Astrophysical IGWs in stars are generally excited by the $l
        = 2$ component of the tidal potential, for which $k_{\perp} \sim
        \frac{1}{R}$ where $R$ is the radius of the star. To best emulate this,
        we use $k_{x} = \frac{2\pi}{L_x}$ the smallest permitted wavenumber
        permitted by periodic boundary conditions.

    \item $\omega$: We choose $\omega$ by evaluating dispersion relation
        $\omega\p{k_{x}, k_{z}}$ for a desired $k_{z}$.

        To choose $k_z$, we note astrophysical IGWs generally satisfy $\omega
        \ll N$, or equivalently $\frac{k_r}{k_{\perp}} \sim k_rR \gg 1$. Since
        $H \lesssim R$, we aim to study $k_{z}H \gg 1$. However, to satisfy this
        well requires high resolution simulations: $L_z \gg H$ is required to
        give waves ample room to grow within the simulation domain, but we want
        $k_{z}$ to be sufficiently separated from the grid spacing $\sim L_z /
        N_z$ that nonlinear effects can be seen in later simulations (see
        \autoref{ss:nl_ns}). Thus, $N_z$ sets how well we can satisfy $k_{z}H
        \gg 1$; we choose $k_{z}H = 1$ as a compromise.

@@

    \item $\nu$: Nonlinear effects transfer wave energy from $\bm{k}$ to
        larger wavenumbers. Since well-resolved simulations using spectral
        methods have no inherent numerical viscosity, energy will accumulate at
        grid scales in the absence of artificial dissipation. We introduce
        dissipation parameter $\nu$ used for both artificial viscosity and
        artificial diffusivity. We ensure our equations of motion continue to
        conserve horizontal momentum (see Appendix~\ref{se:strat_impl} for
        details). Under weak forcing, we may set $\nu = 0$, since energy
        transfer out of the parent $\bm{k}$ mode is negligible.
\end{itemize}

@@

We aim to damp waves that reach the edge of the simulation domain without
inducing nonphysical reflection. To do so, we replace material derivatives
in \autoref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z)\Upsilon,\nonumber\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2,+ \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$; we use $\Delta z = 0.025L_z$. This prescription, used in
similar studies \citep{lecoanet_damp}, has the advantage of being smooth,
important for spectral methods. Further details of our implementation of the
fluid equations in Dedalus are described in Appendix~\ref{se:strat_impl}.

@@

We describe the results of a simulation satisfying the parameter choices in
\autoref{ss:numerics} such that $\xi_z k_z \ll 1$ everywhere in the simulation
domain.

During the simulation, we expect IGW of form \autoref{eq:uz_lin} to be excited.
Invoking incompressibility, we obtain a complete analytical solution for flow
velocities $\bm{u}_{al}'(x, z, t)$. The amplitude of the observed IGW in the
simulation field $\bm{u}$ relative to analytical solution $\bm{u}_{al}'$ over
some region $z \in [z_b, z_t]$ can be estimated with estimator $\hat{A}_i(t)$
(subscript $i$ denotes incident wave)
\begin{equation}
    \hat{A}_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{al}'}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{u_{al}'}^2\;\mathrm{d}x\mathrm{d}z}
        \label{eq:ahat_def}
\end{equation}
If $\bm{u} = \bm{u}_{al}'$, then $\hat{A}_i(t) = 1$. The energy norm is a
traditional normalization choice such that the overlap between $\bm{u},
\bm{u}_{al}' \propto e^{z/2H}$ is evenly weighted throughout the integration
region.

For a simulation satisfying $\xi_z k_z \ll 1$ everywhere, we expect
$\hat{A}_i(t) = 1$ when integrated between the forcing and damping zones,
corresponding to choices $z_b \gtrsim z_0, z_t \lesssim z_T$ ($z_0, z_T$ are
defined in \autoref{eq:vol_drive} and \autoref{eq:Gamma} respectively). For
consistency with the nonlinear case later, we choose $z_b = z_0 + 3\sigma, z_t =
z_b + H$ (using $z_t = z_T - \Delta z$ just below the damping layer instead does
not change the results). The resulting measurement of $\hat{A}_i(t)$ is shown in
\autoref{fig:lin_amps}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_amps.png}
    \caption{Amplitude of excited wave over time in weak forcing simulation,
    computed using \autoref{eq:ahat_def}. $\hat{A}_i(t) = 1$ corresponds to
    perfect agreement with the analytical estimate. After an initial transient
    phase, we observe great agreement with \autoref{eq:uz_lin}, and in
    particular $\hat{A}_i(t)$ asymptotes to a constant value, implying continuous
    excitation of identical IGW.}\label{fig:lin_amps}
\end{figure}

The analytical theory also predicts that the horizontal momentmum flux $F(z, t)$
should be spatially flat between the forcing zone where it is generated and the
damping zone where it is dissipated (\autoref{eq:F_def}). We make may a slightly
more accurate estimate of the flux carried in the IGW by directly substituting
$\bm{u}_{al}'$ into \autoref{eq:F_def}. Calling this estimate
\begin{equation}
    F_{al}'(z) \equiv \ev{\overline{\rho} u_{al,x}'u_{al,z}'}_x
\end{equation}
we may then measure the agreement of our simulation with analytical expectation
by computing
\begin{equation}
    \hat{F}(z, t) \equiv \frac{\ev{\rho u_xu_z}_x}
        {F_{al}'}.\label{eq:sbm_def}
\end{equation}
For our weak forcing simulation, $\hat{F}(z) = 1$ is expected between $z_0, z_T$
after initial transients dissipate, and indeed we observe agreement with this in
\autoref{fig:lin_fluxes}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_fluxes.png}
    \caption{\autoref{eq:sbm_def} as a function of $z$ at select times $t$. As
    initial transients die out, $\hat{F}(z, t) = 1$ to very good agreement above
    the forcing zone $z > z_0 = 2H$ and below the damping zone $z \lesssim z_T =
    9.5H$. The flux excited in the forcing zone is transported without loss to
    the top of the domain, where it is dissipated by the damping layer (see
    \autoref{ss:damping}) without reflection.}\label{fig:lin_fluxes}
\end{figure}

@@

To perform simulations of wave breaking phenomena, we use the same values as
\autoref{ss:numerics} except for $C, \nu$. In particular, we choose $C$ such
that $\at{\xi_z k_z}_{z_0} = 0.1$ in the forcing zone, which implies $\xi_z k_z
$ exceeds $1$ before $z_T$ the upper damping zone. The dissipation parameter
$\nu$ was varied across the various simulations. To quantify $\nu$, we define
dimensionless Reynolds number
\begin{equation}
    \mathrm{Re}  \equiv \frac{\omega}{\nu k_{z}^2}. \label{eq:re_def}
\end{equation}
A table of our simulation viscosities can be found in \autoref{tab:params}.
\begin{table}
    \centering
    \begin{tabular}{l c c c}
        Resolution & $\mathrm{Re}$\\\bottomrule
        $1024 \times 4096$ & $2048$\\
        $768 \times 3072$ & $1024$\\
        $512 \times 2048$ & $512$\\
        $256 \times 1024$ & $341$\\
        $256 \times 1024$ & $205$\\
        $256 \times 1024$ & $146$\\
    \end{tabular}
    \caption{Table of simulation resolutions. Note that $\mathrm{Re}$ is defined
    as in \autoref{eq:re_def}.}\label{tab:params}
\end{table}

@@

A full video of our higher-resolution run at $N_x = 768, N_z = 3072,
\mathrm{Re} = 1024$ is available
online\footnote{http://www.princeton.edu/~lecoanet/data/breaking\_wave.mov}. We
take this to be our fiducial simulation for the remainder of this paper, though
other simulations show qualitatively identical behavior. Snapshots of the
velocity field $\bm{u}$ at select times can be found in \autoref{fig:snapshots}.
% convert yubo_000054.png -crop 2000x2000+150+250 out.png
\begin{figure*}
    \subfloat[$t=413.4$ At early times, the flow resembles a linear IGW lower in
    the simulation domain but breaks down into smaller-scale features at higher
    $z$. Some characteristic swirling motion can be seen in both $u_x, u_z$,
    highly suggestive of Kelvin-Helmholtz instabilities.]{
    \includegraphics[width=0.45\textwidth]{plots/yubo_000054.png}}\hfil
    %
    \subfloat[$t=658.5$ At a slightly later time, the mean flow in $u_x$ has
    become much more prominent and the critical layer $z_c$ has become much more
    definite. Small-scale fluctuations are still present in $u_z$ albeit at
    smaller amplitudes due to being in a denser region of the fluid.]{
    \includegraphics[width=0.45\textwidth]{plots/yubo_000086.png}}

>>>>

    \newcommand*{\scinot}[2]{#1\times10^{#2}}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\mathbf{#1}}
    \newcommand*{\uv}[1]{\hat{\mathbf{#1}}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}
    \DeclareMathOperator*{\erf}{erf}

@@

In practice, it is convenient to introduce the coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$ \citep[e.g.][]{lecoanet_anel}. This both
identically enforces $\rho > 0$ and eliminates the stiff term
$\frac{\bm{\nabla}P}{\rho}$. We also define reduced pressure $\varpi =
\frac{P}{\rho}$. Then, we may rewrite the second two equations of
\autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}}{t} + \bm{\nabla}\varpi + \varpi\bm{\nabla}\Upsilon
            - \frac{\varpi}{H}\uv{z} + g\uv{z} &= 0\label{eq:nl_upsilon_u}.
    \end{align}
\end{subequations}
In the new coordinates, hydrostatic equilibrium corresponds to $\Upsilon = 0,
\overline{\varpi} = gH$.

@@

    \item The averaged horizontal momentum flux $F$ is
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def}\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        Thus, indeed $F > 0$ for an upward propagating IGW $c_{g, z} > 0$.
\end{itemize}

@@

The detailed onset of wave breaking has been laid out in a few key papers
\citep{drazin, klostermeyer, winters1994}, where instabilities transfer momentum
out of the IGW through daughter modes into the mean flow of the fluid. When
viscosity is sufficiently weak (see \autoref{ss:disp} for a discussion of this
assumption in WDs), a non-uniform shear flow and eventually a critical layer
develop. The further evolution of the critical layer with the incident IGW is
thought to be responsible for tidal synchronization in stellar binaries
\citep{zahn75,gn89} as well as terrestrial phenomena such as the quasi-biennial
oscillation \citep{lindzen_qbo}.

The behavior of an IGW incident upon a critical layer was first studied in the
inviscid, linear regime in \citep{booker_bretherton}, which found nearly
complete absorption of the IGW\@. In particular, the incident wave has amplitude
reflection and transmission coefficients that can be expressed in terms of the
local Richardson number Ri at the height of the critical layer $z_c$:
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\overline{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R}_A &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T}_A &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}.
        \label{eq:crit_coeffs}
\end{align}
In the $\mathrm{Ri} \gg 1$ limit, $\mathcal{R}, \mathcal{T} \ll 1$
and the incident wave is completely absorbed to a good approximation.

This result extends to viscous fluids \citep{hazel}, but weakly nonlinear theory
\citep{brown_stewartson} and numerical simulations \citep{winters1994} suggest
that nonlinear effects may significantly enhance reflection and transmission.

Consider the horizontal momentum transfer at the critical layer. Any incident
horizontal momentum flux absorbed by the fluid must manifest as additional
horizontal momentum of the shear flow. Since the mean flow cannot exceed
$\overline{U}_c \equiv \omega/k_x$, the horizontal phase velocity of the
incident wave, the critical layer will instead propagate downward in response to
the incident momentum flux. The horizontal momentum of the shear flow satisfies
\begin{equation}
    \pd{}{t}\int\limits \overline{\rho}(z) \overline{U}(z, t)\;\mathrm{d}z
        - \Delta F = 0,
\end{equation}
where $\Delta F > 0$ is the absorbed horizontal momentum. Assuming
$\overline{U}(z > z_c) = \overline{U}_c, \overline{U}(z < z_c) = 0$, this
condition becomes
\begin{equation}
    \overline{\rho}(z_c) \overline{U}_c\rd{z_c}{t} = \Delta F.\label{eq:zc_anal}
\end{equation}
If $\Delta F$ is constant in time, $z_c(t)$ has analytical solution
\begin{equation}
    z_c(t) = -H\ln \frac{\overline{\rho}(t = 0)
        + t\frac{\Delta F}{\overline{U}_c H}}{\overline{\rho}_0}
        ,\label{eq:zc_sol}
\end{equation}
where $\overline{\rho}(t_0)$ is the density at the critical layer height at $t =
0$.

\section{Internal Gravity Waves: Weakly Forced Numerical
Simulations}\label{s:sim}

@@

We verify the analytic theory of \autoref{s:theory} by running weakly forced
direct numerical simulations using the pseudo-spectral code Dedalus
\citep{dedalus,dedalus2}. IGWs are excited deep within WDs and must propagate
across many density scale heights before reaching the surface. The waves stay
linear until they are within a few scale heights of the breaking height. We thus
simulate only the upper part of the WD\@. Although turbulence driven by wave
breaking is fundamentally 3D \citep{klostermeyer}, we simplify the problem by
restricting our simulations to two dimensions. Future work will study the
problem in 3D spherical geometry.

For numerical convenience, we simulate equations \autoref{eq:nl_incomp},
\autoref{eq:nl_upsilon_u}, and \autoref{eq:vol_drive} in a Cartesian box with
size $L_x, L_z$. The $x$ direction represents the azimuthal direction in a star,
and the $z$ direction represents the radial direction. We thus choose periodic
boundary conditions in the $x$ direction. We also use periodic boundary
conditions in the $z$ direction, but we damp perturbations to zero near the
top/bottom of the domain to capture their distinct behavior (described in
\autoref{ss:damping}). We expand all variables as a Fourier series with $N_x$
and $N_z$ modes, and use the $3/2$ dealiasing rule to avoid aliasing errors in
the nonlinear terms \citep{boyd}.

The geometry of our simulation domain is fixed by one further parameter: $z_0$
the forcing location. We choose $L_z = 10H$ to give $\sim e^3$ amplitude growth
between the damping zones. We force at $z_0 = 2H = 0.2L_z$ within width
$\sigma = 0.078H$, sufficiently far from the lower damping zone and permit
sufficient room for the upward moving wave to grow as $\propto e^{z/2H}$.
Finally, we want similar grid spacing $\frac{L_x}{N_x} \sim \frac{L_z}{N_z}$,
guided by the intuition that turbulence is approximately isotropic, so we use
$L_x = 4H$ and $N_z / N_x = 4$.

@@

We non-dimensionalize the problem such that $H = N = \rho_0 = 1$. The physics of
the simulation is then fixed by the four remaining parameters $k_{x}, \omega, C,
\nu$. We describe our choices for these parameters below:
\begin{itemize}
    \item $k_{x}$: Tidally excited waves in stars generally have $\ell = 2$,
        corresponding to a very small $k_\perp\sim 1/R$, where $R$ is the radius
        of the star. We use the smallest wavenumber in our simulation,
        $k_x=2\pi/L_x$.

    \item $\omega$: We choose $\omega$ by evaluating the dispersion relation
        $\omega(k_x, k_z)$ for a desired $k_z$. We pick $\abs{k_z H} = 2\pi$ to
        ensure the waves are very well resolved in all of our simulations. Note
        however that tidally forced IGWs typically have $\omega \ll N$, or
        equivalently $k_r/k_\perp \sim k_r R \gg 1$. This requires $k_z H
        \gtrsim 1$, which is only marginally satisfied in our simulations.

@@

    \item $\nu$: Nonlinear effects transfer wave energy from the injection
        wavenumber $\bm{k}$ to larger wavenumbers. Our spectral method does not
        have any numerical viscosity, so diffusivity must be introduced into the
        equations to regularize the systems. We add viscosity and diffusivity to
        the system in a way that conserves horizontal momentum (see
        Appendix~\ref{se:strat_impl} for details). When the forcing is weak, we
        may set $\nu=0$ because there is negligible energy transfer between
        modes.
\end{itemize}

@@

We aim to damp waves that reach the edge of the simulation domain without
inducing nonphysical reflection. To do so, we replace material derivatives
in \autoref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z)\Upsilon,\nonumber\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$. We use $\Delta z = 0.025L_z$ and $\tau = 1/15N$. This
prescription is similar to \citep{lecoanet_damp} and has the advantage of being
smooth, important for spectral methods. Further details of our implementation of
the fluid equations in Dedalus are described in Appendix~\ref{se:strat_impl}.

@@

We describe the results of a simulation with $C = \scinot{1.64}{-7}$ such that
$\xi_z k_z \ll 1$ everywhere. In particular, $\xi_z k_z \approx \scinot{5}{-5}$
just above the forcing zone.

We expect the waves to follow the solution given by \autoref{eq:uz_lin}. Along
with incompressibility (\autoref{eq:nl_incomp}), we obtain a complete analytical
solution for flow velocities $\bm{u}_{al}(x, z, t)$. The amplitude of the
observed IGW in the simulation field $\bm{u}$ relative to analytical solution
$\bm{u}_{al}$ over some region $z \in [z_b, z_t]$ can be estimated with
estimator $\hat{A}_i(t)$ (subscript $i$ denotes incident wave)
\begin{equation}
    \hat{A}_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{al}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\abs{\bm{u}_{al}}^2\;\mathrm{d}x\mathrm{d}z}
        \label{eq:ahat_def}
\end{equation}
If $\bm{u} = \bm{u}_{al}$, then $\hat{A}_i(t) = 1$. The energy norm is a
traditional normalization choice such that the overlap between $\bm{u},
\bm{u}_{al} \propto e^{z/2H}$ is evenly weighted throughout the integration
region.

For a simulation satisfying $\xi_z k_z \ll 1$ everywhere, we expect
$\hat{A}_i(t) = 1$ when integrated between the forcing and damping zones, i.e.\
$z_b \gtrsim z_0, z_t \lesssim z_T$ ($z_0, z_T$ are defined in
\autoref{eq:vol_drive} and \autoref{eq:Gamma} respectively). For consistency
with the nonlinear case later, we choose $z_b = z_0 + 3\sigma, z_t = z_b + H$
(using $z_t = z_T - \Delta z$ just below the damping layer instead does not
change the results). The resulting measurement of $\hat{A}_i(t)$ is shown in
\autoref{fig:lin_amps}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_amps.png}
    \caption{Amplitude of excited wave over time in weak forcing simulation,
    computed using \autoref{eq:ahat_def}. $\hat{A}_i(t) = 1$ corresponds to
    perfect agreement with the analytical estimate. After an initial transient
    phase, we observe great agreement with \autoref{eq:uz_lin}, and in
    particular $\hat{A}_i(t)$ asymptotes to $\approx 1$, implying continuous
    excitation of identical IGW with the expected
    amplitude.}\label{fig:lin_amps}
\end{figure}

The analytical theory also predicts that the horizontal momentmum flux $F(z, t)$
should be independent of $z$ between the forcing zone where it is generated and
the damping zone where it is dissipated (\autoref{eq:F_def}). We make may a
slightly more accurate estimate of the flux carried in the IGW by directly
substituting $\bm{u}_{al}$ into \autoref{eq:F_def}. Calling this estimate
\begin{equation}
    F_{al}'(z) \equiv \ev{\overline{\rho} u_{al,x}u_{al,z}}_x
\end{equation}
we may then measure the agreement of our simulation with analytical expectation
by computing
\begin{equation}
    \hat{F}(z, t) \equiv \frac{\ev{\rho u_xu_z}_x}
        {F_{al}'}.\label{eq:sbm_def}
\end{equation}
\autoref{fig:lin_fluxes} shows that $\hat{F}(z)\approx 1$ between $z_0$ and
$z_T$, as expected.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_fluxes.png}
    \caption{\autoref{eq:sbm_def} as a function of $z$ at select times $t$. As
    initial transients die out, $\hat{F}(z, t) = 1$ to very good agreement above
    the forcing zone $z > z_0 = 2H$ and below the damping zone $z \lesssim z_T =
    9.5H$. The flux excited in the forcing zone is transported without loss to
    the top of the domain, where it is dissipated by the damping layer (see
    \autoref{ss:damping}) without reflection.}\label{fig:lin_fluxes}
\end{figure}

@@

To perform simulations of wave breaking phenomena, we use the same values as
\autoref{ss:numerics} except for $C$ and $\nu$. In particular, we choose $C$
such that $\at{\xi_z k_z}_{z_0} = 0.1$ in the forcing zone. The linear solution
predicts $\xi_z k_z \sim 4.25$ at the upper damping zone $z_T$. The dissipation
parameter $\nu$ was varied across the various simulations. To quantify $\nu$, we
define dimensionless Reynolds number
\begin{equation}
    \mathrm{Re}  \equiv \frac{\omega}{\nu k_{z}^2}. \label{eq:re_def}
\end{equation}
A table of our simulations can be found in \autoref{tab:params}.
\begin{table}
    \centering
    \begin{tabular}{l c c c}
        Resolution & $\mathrm{Re}$\\\bottomrule
        $1024 \times 4096$ & $2048$\\
        $768 \times 3072$ & $1024$\\
        $512 \times 2048$ & $512$\\
        $256 \times 1024$ & $341$\\
        $256 \times 1024$ & $205$\\
        $256 \times 1024$ & $146$\\
    \end{tabular}
    \caption{Table of simulation resolutions.}\label{tab:params}
\end{table}

@@

A full video of our higher-resolution run at $N_x = 768$, $N_z = 3072$,
$\mathrm{Re} = 1024$ is available
online\footnote{http://www.princeton.edu/~lecoanet/data/breaking\_wave.mov}. We
take this to be our fiducial simulation for the remainder of this paper, though
other simulations show qualitatively similar behavior.

% convert yubo_000054.png -crop 2000x2000+150+250 out.png
In \autoref{fig:snapshots}, we present snapshots of the velocity field $\bm{u}$
at various interesting phases of the simulation.
\begin{enumerate}
    \item At early times (top left panel), the flow resembles a linear IGW lower
        in the simulation domain but breaks down into smaller-scale features at
        higher $z$. Some characteristic swirling motion can be seen in both
        $u_x, u_z$, highly suggestive of Kelvin-Helmholtz instabilities.

    \item At a slightly later time (top right panel), the mean flow in
        $u_x$ has become much more prominent and the critical layer $z_c$ has
        become much more definite. Small-scale fluctuations are still present in
        $u_z$ but at smaller amplitudes due to being in a denser region of the
        fluid.

    \item In the bottom left panel, the critical layer transition is now
        extremely sharp, and small swirls of limited vertical extent at the
        location of the critical layer in $u_z$ suggest that the
        Kelvin-Helmholtz instability is responsible for regulating the minimum
        width of this transition. This is supported by the conclusions in
        \autoref{ss:khi}.

    \item  The end of the simulation shows very few significant qualitative
        differences from the previous snapshot, suggesting that the latter half
        of our simulation is temporally converged.
\end{enumerate}

\begin{figure*}
    \includegraphics[width=0.45\textwidth]{plots/yubo_000054.png}\hfil
    %
    \includegraphics[width=0.45\textwidth]{plots/yubo_000086.png}
